---
title: "Homework 6 - Statistical Data Analysis"
author: "Yigit Kasal"
date: "2025-06-02"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(readr)
library(tidyverse)
library(Rmisc)
library(languageR)
library(car)
library(lingpsych)
library(lmerTest)
```

# Exercise 1

## Load and prepare data

```{r}
stress <- read_delim("StressSymptoms2.txt",
                     delim = "\t",
                     locale = locale(encoding = "UTF-8"),
                     trim_ws = TRUE)

stress$gender_coded <- ifelse(stress$gender == "man", -0.5,
                              ifelse(stress$gender == "woman", 0.5, NA_real_))
```

## Fit regression with interaction

```{r}
mod <- lm(symptoms ~ stress * gender_coded, data = stress)
summary(mod)$coefficients
```

**Interpretation:**  
Intercept = 71.047  
Slope (stress) = 0.91  
Slope (gender_coded) = -16.77  
Interaction = 0.85

## Manually add interaction term

```{r}
stress$stressXgender <- stress$stress * stress$gender_coded

mod2 <- lm(symptoms ~ stress + gender_coded + stressXgender, data = stress)
summary(mod2)$coefficients
```

**Observation:**  
Manually creating the interaction term gives the same results as using `*`.

# Exercise 2

## Load and prepare EEG data

```{r}
mydata <- read.csv("EEG_Indiv_RT_Dataset20180706out.csv", sep = ";")
mydata <- mydata[mydata$Accuracy == "1", ]
mydata <- mydata[!is.na(mydata$Latency), ]
mydata <- mydata[!is.na(mydata$AA), ]
mydata$AoA <- as.factor(as.character(mydata$AoA))
contrasts(mydata$AoA) <- c(-0.5, 0.5)
```

## Check multicollinearity

```{r}
ex2m <- lmer(Latency ~ AoA + freqfilms2 + (1 | PP.id) + (1 | Item), data = mydata, REML = FALSE)
summary(ex2m)

lm_vif <- lm(Latency ~ AoA + freqfilms2, data = mydata)
car::vif(lm_vif)
```

## Linear model with both predictors

```{r}
lm_full <- lm(Latency ~ freqfilms2 + AoA, data = mydata)
summary(lm_full)$coefficients
```

## Likelihood ratio test

```{r}
m_freq <- lmer(Latency ~ freqfilms2 + (1 | PP.id) + (1 | Item), data = mydata, REML = FALSE)
m_full <- lmer(Latency ~ freqfilms2 + AoA + (1 | PP.id) + (1 | Item), data = mydata, REML = FALSE)
anova(m_freq, m_full)
```

**Conclusion:**  
Even after controlling for frequency, AoA significantly affects naming latency.

## Interaction effect

```{r}
mydata$freq_c <- with(mydata, freqfilms2 - mean(freqfilms2, na.rm = TRUE))

m_int <- lmer(Latency ~ AoA * freq_c + (1 | PP.id) + (1 | Item), data = mydata, REML = FALSE)
summary(m_int)$coefficients
```

**Conclusion:**  
No significant interaction between AoA and frequency. AoA has a consistent effect.

## Assumption checks

```{r}
mydata$invLatency <- 1 / mydata$Latency
m_inv <- lmer(invLatency ~ AoA * freq_c + (1 | PP.id) + (1 | Item), data = mydata, REML = FALSE)

par(mfrow = c(1, 2))
qqnorm(resid(m_int), main = "m_int QQ-plot (raw latency)")
qqline(resid(m_int))

plot(fitted(m_int), resid(m_int), xlab = "Fitted values", ylab = "Residuals",
     main = "m_int Residuals vs fitted")
abline(h = 0, lty = 2)

par(mfrow = c(1, 2))
qqnorm(resid(m_inv), main = "m_inv QQ-plot (1/latency)")
qqline(resid(m_inv))

plot(fitted(m_inv), resid(m_inv), xlab = "Fitted values", ylab = "Residuals",
     main = "m_inv Residuals vs fitted")
abline(h = 0, lty = 2)

par(mfrow = c(1, 1))
```

**Interpretation:**  
Reciprocal transformation improves normality and homoscedasticity of residuals.
